Reducing RES Droughts through the integration of wind and PV

### 2025-01-09 Article/main.tex
Update Article

The spreadsheet available from the EirGrid website contains two key variables: generation and availability. Generation is the energy that a RES farm actually contributed to the grid, which may include limitations introduced by the TSO to maintain grid stability, such as constraints and curtailment. Availability represents the energy that would have been generated from a RES farm if no grid constraints had been applied, making it representative of the weather-related response. Generation and availability values are available from 2014 onward for wind power and from 2018 onward for PV power, although PV only became present in the Republic of Ireland data in 2023. This study focuses on availability for all analyses.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present at a horizontal resolution of 0.25\textdegree, which is approximately every 27\,km. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables used by Atlite and C3S-Energy.

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 meters. This choice is based on expert advice and the trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: Surface solar radiation downwards ($ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times, using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

Fig.~\ref{fig:bar_number_events_verification_wind} presents the average annual number of wind drought events during the 2014 to 2023 validation period. The figure shows that Atlite shows the best agreement overall with the observed frequency and duration of wind drought events. This pattern is particularly evident for shorter-duration events, which are the most frequent.

@@

\begin{table}[!ht]
	\centering
	\begin{tabular}{l|lll|}
	\cline{2-4}
	& \textbf{Atlite} & \textbf{C3S-E G} & \textbf{C3S-E N} \\ \hline
	\multicolumn{1}{|l|}{\textbf{CC}}   & 0.921           & 0.931            & 0.931            \\ \hline
	\multicolumn{1}{|l|}{\textbf{RMSE}} & 0.119           & 0.090            & 0.113            \\ \hline
	\multicolumn{1}{|l|}{\textbf{MBE}}   & 0.046           & 0.027           & 0.021           \\ \hline
	\end{tabular}
	\caption{Skill scores for wind CF for the three datasets compared to observed data}
	\label{tab:pv_skill_scores}
\end{table}

@@

Nevertheless, the ultimate goal of this analysis is to assess the combination of wind and PV generation, where the complementary nature of these energy sources mitigates the limitations seen in PV-only results.

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets. For Fig.~\ref{fig:boxplot_number_events}a, the number of events decreases as the duration range increases, with very few events lasting more than seven days. For Fig.~\ref{fig:boxplot_number_events}b, the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a~\&~b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV power compared to wind power, especially in a region such as Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

@@

For the existing installed capacity scenario (Fig.\ref{fig:return_periods}c), the return periods mirror those of Fig.\ref{fig:return_periods}a, due to the low levels of installed PV capacity. In the balanced scenario (Fig.~\ref{fig:return_periods}d), the return periods for RES droughts increase across all durations. For example, the return period for a five-day drought event, shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}, extends from roughly six months for the existing scenario, to four years for the balanced scenario in the Atlite dataset, and from about fifteen months to around five years in the two C3S-E datasets.

@@

Across Fig.\ref{fig:return_periods}a, c, and, d, the return periods in the Atlite dataset are consistently higher than those in the two C3S-E datasets. For instance, in the existing scenario (Fig.\ref{fig:return_periods}c), an event with a one-year return period lasts six days in the Atlite dataset, compared to only five days in the C3S-E datasets. This difference underscores the importance of model selection when quantifying RES droughts, as each model’s assumptions and parametrisations significantly influence drought duration estimates. Additionally, in all four graphs, the similarity between results from the two C3S-E datasets suggests that assumptions in the Atlite model—such as wind turbine power curve selection and PV panel specifications—have a greater impact on RES drought duration estimates than the precise geographic distribution of RES farms when studying the return periods of RES droughts.

@@

Similar to previous results, the existing installed capacity wind-PV combination (Fig.\ref{fig:res_droughts_seasonality}c) shows patterns comparable to \ref{fig:res_droughts_seasonality}a. However, in this panel, the number of hours classified as RES droughts in summer decreases slightly compared to the wind-only scenario. This reduction can be explained by the contribution of PV generation during the summer months in the existing scenario, even though it constitutes only 11\% of total capacity. Since the number of RES drought hours for PV in summer is near zero, this small contribution has a noticeable impact on reducing overall drought hours. In the balanced scenario (Fig.~\ref{fig:res_droughts_seasonality}d), all three datasets show a reduction in monthly RES drought frequency. Annual reductions in median RES drought frequency are observed across the datasets, dropping from 14\% to 5\% for Atlite, from 8\% to 3\% for C3S-E G, and from 9\% to 4\% for C3S-E N. The balanced mix of wind and PV power in this scenario reduces the seasonal signal overall and significantly decreases the percentage of RES drought hours in the summer. By reducing the frequency of summer droughts while maintaining resilience during winter, the balanced scenario offers a more reliable energy supply that better aligns with seasonal demand patterns in Ireland.

@@

Atlite shows the best alignment with observed data for wind generation. Differences between the models are smaller for PV, with C3S-G performing marginally better than the other two. The results show that the two C3S-E datasets (C3S-E G and C3S-E N) consistently yield similar outcomes, indicating that their methodological differences have minimal impact. This distinction was also evident in the analysis, where Atlite reported higher return periods and a greater number of RES droughts, especially in scenarios with a balanced share of RES. Again, the results from RES drought modelling relies more on the precision of the wind power curve and PV panel models than on the specific locations of RES farms. Atlite’s superior performance highlights the importance of selecting validated models for assessing RES drought risks. This careful model selection can better quantify risks, support effective planning, and avoid the potential underestimation of capacity needs, which is essential for ensuring energy security.

Looking at the balanced scenario, the analysis showed a significant improvement in the management of RES droughts due to the complementary nature of wind and PV generation. Wind and PV together perform better in terms of reducing drought frequency and duration than either would individually, largely because of the seasonal anti-correlation between the two energy sources. This diversification reduces the seasonal impact on RES droughts, as PV generation peaks in the summer and wind generation is more consistent in winter. Ireland currently has a highly wind-dependent energy system, but with ambitious targets for PV installations in the coming years, the energy mix is expected to approach a balance between wind and PV capacity. While this balanced approach offers a more stable and secure energy supply by mitigating RES drought risks, it is important to note that having similar wind and PV capacities may not optimize other aspects, such as annual energy production or meeting nighttime loads. For policymakers, these findings underscore the importance of meeting these capacity targets to enhance energy security through diversification. Additionally, the choice of model for RES drought assessment becomes increasingly critical as more renewable capacity is integrated into the system.

>>>>

The spreadsheet available from the EirGrid website contains two key variables: generation and availability. Generation is the energy that a RES farm actually contributed to the grid, which may include limitations introduced by the TSO to maintain grid stability, such as constraints and curtailment. Availability represents the energy that would have been generated from a RES farm if no grid constraints had been applied, making it representative of the weather-related response. Generation and availability values are available from 2014 onward for wind power and from 2018 onward for PV power, although PV availability data only became present in the Republic of Ireland in 2023. This study focuses on availability for all analyses.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present at a horizontal resolution of 0.25\textdegree. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables used by Atlite and C3S-Energy.

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 meters. This choice is based on expert advice and the trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: surface solar radiation downwards ($ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times, using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

Fig.~\ref{fig:bar_number_events_verification_wind} presents the average annual number of wind drought events during the 2014 to 2023 validation period. The figure presents that Atlite shows the best agreement overall with the observed frequency and duration of wind drought events. This pattern is particularly evident for shorter-duration events, which are the most frequent.

@@

\begin{table}[!ht]
	\centering
	\begin{tabular}{l|lll|}
	\cline{2-4}
	& \textbf{Atlite} & \textbf{C3S-E G} & \textbf{C3S-E N} \\ \hline
	\multicolumn{1}{|l|}{\textbf{CC}}   & 0.921           & 0.931            & 0.931            \\ \hline
	\multicolumn{1}{|l|}{\textbf{RMSE}} & 0.119           & 0.090            & 0.113            \\ \hline
	\multicolumn{1}{|l|}{\textbf{MBE}}   & 0.046           & 0.027           & 0.021           \\ \hline
	\end{tabular}
	\caption{Skill scores for PV CF for the three datasets compared to observed data}
	\label{tab:pv_skill_scores}
\end{table}

@@

Nevertheless, the goal of this analysis is to assess the combination of wind and PV generation, where the complementary nature of these energy sources mitigates the limitations seen in PV-only results.

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets. For Fig.~\ref{fig:boxplot_number_events}a, the number of events decreases as the duration range increases, with very few events lasting more than seven days. For Fig.~\ref{fig:boxplot_number_events}b, the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a~\&~b), the median, first, and third quartiles for PV are consistently higher than or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV power compared to wind power, especially in a region such as Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

@@

For the existing installed capacity scenario (Fig.~\ref{fig:return_periods}c), the return periods mirror those of Fig.~\ref{fig:return_periods}a, due to the low levels of installed PV capacity. In the balanced scenario (Fig.~\ref{fig:return_periods}d), the return periods for RES droughts increase across all durations. For example, the return period for a five-day drought event, shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}, extends from roughly six months for the existing scenario, to four years for the balanced scenario in the Atlite dataset, and from about fifteen months to around five years in the two C3S-E datasets.

@@

Across Fig.~\ref{fig:return_periods}a, c, and, d, the return periods in the Atlite dataset are consistently higher than those in the two C3S-E datasets. For instance, in the existing scenario (Fig.~\ref{fig:return_periods}c), an event with a one-year return period lasts six days in the Atlite dataset, compared to only five days in the C3S-E datasets. This difference underscores the importance of model selection when quantifying RES droughts, as each model’s assumptions and parametrisations significantly influence drought duration estimates. Additionally, in all four graphs, the similarity between results from the two C3S-E datasets suggests that assumptions in the Atlite model—such as wind turbine power curve selection and PV panel specifications—have a greater impact on RES drought duration estimates than the precise geographic distribution of RES farms when studying the return periods of RES droughts.

@@

Similar to previous results, the existing installed capacity wind-PV combination (Fig.~\ref{fig:res_droughts_seasonality}c) shows patterns comparable to \ref{fig:res_droughts_seasonality}a. However, in the 91W/9PV scenario, the number of hours classified as RES droughts in summer decreases slightly compared to the wind-only scenario. This reduction can be explained by the contribution of PV generation during the summer months in the existing scenario, even though it constitutes only 11\% of total capacity. Since the number of RES drought hours for PV in summer is near zero, this small contribution has a noticeable impact on reducing overall drought hours. In the balanced scenario (Fig.~\ref{fig:res_droughts_seasonality}d), all three datasets show a reduction in monthly RES drought frequency. Annual reductions in median RES drought frequency are observed across the datasets, dropping from 14\% to 5\% for Atlite, from 8\% to 3\% for C3S-E G, and from 9\% to 4\% for C3S-E N. The balanced mix of wind and PV power in this scenario reduces the seasonal signal overall and significantly decreases the percentage of RES drought hours in the summer. By reducing the frequency of summer droughts while maintaining resilience during winter, the balanced scenario offers a more reliable energy supply that better aligns with seasonal demand patterns in Ireland.

@@

Atlite shows the best alignment with observed data for wind generation. Differences between the models are smaller for PV, with C3S-G performing marginally better than the other two. The results show that the two C3S-E datasets (C3S-E G and C3S-E N) consistently yield similar outcomes, indicating that their methodological differences have minimal impact. This distinction was also evident in the analysis, where Atlite reported higher return periods and a greater number of RES droughts, especially in scenarios with a balanced share of RES. Again, the results from RES drought modelling rely more on the precision of the wind power curve and PV panel models than on the specific locations of RES farms. Atlite’s superior performance highlights the importance of selecting validated models for assessing RES drought risks. This careful model selection can better quantify risks, support effective planning, and avoid the potential underestimation of capacity needs, which is essential for ensuring energy security.

Looking at the balanced scenario, the analysis showed a significant improvement in the management of RES droughts due to the complementary nature of wind and PV generation. Wind and PV together perform better in terms of reducing drought frequency and duration than either would individually, largely because of the seasonal anti-correlation between the two energy sources. This diversification reduces the seasonal impact on RES droughts, as PV generation peaks in the summer and wind generation is more consistent in winter. Ireland currently has a highly wind-dependent energy system, but with ambitious targets for PV installations in the coming years, the energy mix is expected to approach a balance between wind and PV capacity. While this balanced approach offers a more stable and secure energy supply by mitigating RES drought risks, it is important to note that having similar wind and PV capacities may not optimise other aspects, such as annual energy production or meeting nighttime loads. For policymakers, these findings underscore the importance of meeting these capacity targets to enhance energy security through diversification. Additionally, the choice of model for RES drought assessment becomes increasingly critical as more renewable capacity is integrated into the system.
